Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-22 model.tex
Liz is done rewriting

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the predicted value $p_i$ at time $i$ is
calculated using the relation $$p_i = x_{i-1}$$ The prediction
strategy that we refer to using the term ``\naive'' averages the prior
observations to generate the forecast: $$p_i =
\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
simplistic, they are not without merit.  For a time series near the
high end of the complexity spectrum---i.e., one that possesses very
little predictive structure---these two methods can actually be the
best choice.  In forecasting currency exchange rates, for instance,
sophisticated econometrics-based prediction models fail to
consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
These signals are constantly changing, noisy, and possess very little predictive
structure, but their variations are not---on the average---very large, so
the random-walk method's strategy of simply guessing the last known
value is not a bad choice.  If a signal has a unimodal distribution
with low variance, the \naive ~prediction strategy will perform quite
well---even if the signal is highly complex---simply because the mean
is a good approximation of the future behavior.  Moreover, the \naive
~prediction strategy's temporal average effects a low-pass filtering
operation, which effectively mitigates the complexity in signals with
very little predictive structure.

Both of these methods have significant weaknesses, however.  Because
they do not model the temporal patterns in the data, or even the
distribution of its values, they cannot track changes in that
structure.  This causes them to fail in a number of important
situations.  Random-walk strategies are a particularly bad choice for
time series that change significantly at every time step.  In the
worst case---a large-amplitude square wave whose period is equivalent
to twice the sample time---a random-walk prediction would be exactly
180 degrees out of phase with the true continuation.  The \naive
~method would be a better choice in this situation, since it would
always predict the mean.  It would, however, perform poorly when a
signal has a number of long-lived regimes that have significantly
different means.  In this situation, the inertia of the \naive
~method's accumulating mean is a liability and the agility of the
random-walk method is an advantage, since it can respond quickly to
regime shifts.

Of course, methods that could capture and exploit the geometry of the
data, or its temporal patterns, would be far more effective in the
situations described in the previous paragraph.  The ARIMA and LMA
methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
designed to do exactly that. 
% Note, though, that periodic patterns
%appear only in signals that are at the low end of the %complexity
%spectrum. 
However, if a signal contains little predictive structure, forecast
strategies like ARIMA and LMA have nothing to work with and thus will
often be outperformed by the two simple strategies described in this
section.  This is explored further in Sections~\ref{sec:accuracy}
and~\ref{sec:results}.

@@

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that to
make predictions.  The roots of this approach date back to the
original autoregressive schema~\cite{weigend93}, which forecasts the
next time step through a weighted average of past observations: $$p_i
= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
computed generally using either an ordinary least squares approach, or with the method of moments using the Yule-Walker equations.  To account for noise in the data, one can add
a so-called ``moving average'' term to the model; to remove
nonstationarities, one can detrend the data using a differencing
operation.  A strategy that incorporates all three of these features
is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
structure is present in the data, a \emph{seasonal ARIMA model}, which
adds a sampling operation that filters out periodicities, can be a
good choice.

@@

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts can be inaccurate.
Another weakness of this method is prediction horizon: an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To sidestep this issue, we
build forecasts in a stepwise fashion: i.e., fit the model to the
existing data, use it to perform a one-step prediction, rebuild the
model using the latest observations, and iterate until the desired
prediction horizon is reached.  (For consistency, we take the same
approach with the other three models in this study as well, even
though this amounts to artificially hobbling LMA.)

@@

Delay-coordinate embedding~\cite{packard80,Sauer:1991lr,Takens:1981uq}
allows one to reconstruct a dynamical system's full state-space
dynamics from a scalar time-series measurement---provided that some
conditions hold regarding that data.  Specifically, if the underlying
dynamics and the measurement function---the mapping from the unknown
state vector $\vec{X}$ to the observed value $x_i$---are both smooth
and generic, Takens~\cite{Takens:1981uq} formally proves that the
delay-coordinate map
\[
F(\tau,m)(\vec{X}) = ([x_{i} ~ x_{i+\tau} ~ \dots ~x_{i+m\tau}])
\]
from a $d$-dimensional smooth compact manifold $M$ to
$\mathbb{R}^{2d+1}$ is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.  This is an extremely powerful result: among other
things, it means that one can model the full system dynamics without
measuring, or even knowing, every one of its state variables.

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as an
estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.  Plots of
the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored},
embedded following this procedure, are shown in
Figure~\ref{fig:embedding}.
 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

@@

Geometric structure in these kinds of plots is an indication of
structure in the information generation/transmission process that
produced the time series.  The dynamical systems community has
developed a number of methods that leverage this structure to generate
predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
One of the most straightforward of these is the ``Lorenz method of
analogues'' (LMA), which is essentially nearest-neighbor prediction in
the embedded\footnote{Lorenz's original formulation used the full
  system state space; According to \cite{kantz97} this method was first extended to embedded dynamics by Pikovsky ~\cite{pikovsky86-sov}, but more closely related to the prediction work of Sugihara \& May in \cite{sugihara90}}
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
prediction---provides accurate forecasts when the generating process
is a deterministic dynamical system.

Since LMA does not rest on an assumption of linearity (as ARIMA does),
it can handle both linear and nonlinear processes.  If the underlying
generating process is nondeterministic, however, it can perform
poorly.  In Figure~\ref{fig:embedding}(b), for instance, little
structure is visible, so LMA may not work well.  More structure
appears to be present in~Figure~\ref{fig:embedding}(c), but this
reconstruction also appears to contain some noise.  The question as to
how much structure is present in a reconstruction---and how much of
that structure can be captured and used by LMA---is apropos of the
central question treated in this paper.  It may be that \gcc has some
redundancy that LMA cannot exploit, or that the structure in \svdfive
is effectively obfuscated, from the standpoint of the LMA method, by
noise.  By quantifying the balance between normalized (that
is, predictive structure) and entropy for these real-valued time
series, as shown in Section~\ref{sec:results}, we can begin to answer
these questions.

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 360
traces described in Section~\ref{sec:methods}, then calculate the
error between the predictions and the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, which can generate arbitrary-length
predictions---although the accuracy of these predictions will
  degrade over prediction length, in the presence of positive Lyapunov
  exponents, but the performance of the other three methods used here
will degrade severely if the associated models are not periodically
rebuilt.  In order to make the comparison fair, we used an iterative
one-step prediction schema {\sl for all four methods}.  This has the
slightly confusing effect of causing the ``test'' signal to be used
both to assess the accuracy of each model and for periodic refitting.

@@

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:svd5LMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  The shapes of
Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are reminiscent of the
projected embedding in Figure~\ref{fig:embedding}(a).  For a
random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal ordering of the
  points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of a
  projected embedding of the time series, however.}.

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error (MASE) between the true and predicted signals:
% 
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
% 
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  MASE is a normalized measure; the
scaling term in the denominator
% 
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
% 
is the average in-sample forecast error for a random-walk prediction
over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$ means that the
prediction error in question was, on the average, smaller than the
error of a random-walk forecast on the same data.  Analogously,
$MASE>1$ means that the corresponding prediction method did
\emph{worse}, on average, than the random-walk method.  We chose this
error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales.

>>>>

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the predicted value $p_i$ at time $i$ is
calculated using the following relation: $$p_i = x_{i-1}$$ The
prediction strategy that we refer to using the term ``\naive''
averages the prior observations to generate the forecast: $$p_i =
\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
simplistic, they are not without merit.  For a time series near the
high end of the complexity spectrum---i.e., one that possesses very
little predictive structure---these two methods can actually be the
best choice.  In forecasting currency exchange rates, for instance,
sophisticated econometrics-based prediction models fail to
consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
These signals are constantly changing, noisy, and possess very little
predictive structure, but their variations are not---on the
average---very large, so the random-walk method's strategy of simply
guessing the last known value is not a bad choice.  If a signal has a
unimodal distribution with low variance, the \naive ~prediction
strategy will perform quite well---even if the signal is highly
complex---simply because the mean is a good approximation of the
future behavior.  Moreover, the \naive ~prediction strategy's temporal
average effects a low-pass filtering operation, which can  mitigate the
complexity in signals with very little predictive structure.

Both of these methods have significant weaknesses, however.  Because
they do not model the temporal patterns in the data, or even the
distribution of its values, they cannot track changes in that
structure.  This causes them to fail in a number of important
situations.  Random-walk strategies are a particularly bad choice for
time series that change significantly at every time step.  In the
worst case---a large-amplitude square wave whose period is equivalent
to twice the sample time---a random-walk prediction would be exactly
180 degrees out of phase with the true continuation.  The \naive
~method would be a better choice in this situation, since it would
always split the difference.  It would, however, perform poorly when a
signal has a number of long-lived regimes that have significantly
different means.  In this situation, the inertia of the \naive
~method's accumulating mean is a liability and the agility of the
random-walk method is an advantage, since it can respond quickly to
regime shifts.

Of course, methods that could capture and exploit the geometry of the
data, or its temporal patterns, would be far more effective in the
situations described in the previous paragraph.  The ARIMA and LMA
methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
designed to do exactly that. 
% Note, though, that periodic patterns
%appear only in signals that are at the low end of the %complexity
%spectrum. 
However, if a signal contains little predictive structure, forecast
strategies like ARIMA and LMA have nothing to work with and thus will
often be outperformed by the two simple strategies described in this
section.  This effect is explored further in Sections~\ref{sec:accuracy}
and~\ref{sec:results}.

@@

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use it to make
predictions.  The roots of this approach date back to the original
autoregressive schema~\cite{weigend93}, which forecasts the next time
step through a weighted average of past observations: $$p_i =
\sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
generally computed using either an ordinary least squares approach, or
with the method of moments using the Yule-Walker equations.  To
account for noise in the data, one can add a so-called ``moving
average'' term to the model; to remove nonstationarities, one can
detrend the data using a differencing operation.  A strategy that
incorporates all three of these features is called a \emph{nonseasonal
  ARIMA model}.  If evidence of periodic structure is present in the
data, a \emph{seasonal ARIMA model}, which adds a sampling operation
that filters out periodicities, can be a good choice.

@@

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts can be inaccurate.
Another weakness of this method is prediction horizon: an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To sidestep this issue, we
build forecasts in a stepwise fashion: i.e., fit the model to the
existing data, use that model to perform a one-step prediction, rebuild it
using the latest observations, and iterate until the desired
prediction horizon is reached.  (For consistency, we take the same
approach with the other three models in this study as well, even
though doing so amounts to artificially hobbling LMA.)

@@

Delay-coordinate embedding~\cite{packard80,Sauer:1991lr,Takens:1981uq}
allows one to reconstruct a dynamical system's full state-space
dynamics from a scalar time-series measurement---provided that some
conditions hold regarding those data.  Specifically, if the underlying
dynamics and the measurement function---the mapping from the unknown
state vector $\vec{X}$ to the observed value $x_i$---are both smooth
and generic, Takens~\cite{Takens:1981uq} formally proves that the
delay-coordinate map
\[
F(\tau,m)(\vec{X}) = ([x_{i} ~ x_{i+\tau} ~ \dots ~x_{i+m\tau}])
\]
from a $d$-dimensional smooth compact manifold $M$ to
$\mathbb{R}^{2d+1}$ is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.  This is an extremely powerful result: among other
things, it means that one can model the full system dynamics, up to
diffeomorphism, without measuring---or even knowing---every one of its
state variables.

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as
an estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.
Some example plots of data from
Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following
this procedure, are shown in Figure~\ref{fig:embedding}.
 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

@@

Geometric structure in these kinds of plots is an indication of
structure in the information generation/transmission process that
produced the time series.  The dynamical systems community has
developed a number of methods that leverage this structure to generate
predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
One of the most straightforward of these is the \emph{Lorenz method of
  analogues} (LMA), which is essentially nearest-neighbor prediction
in the embedded\footnote{Lorenz's original formulation used the full
  system state space; According to \cite{kantz97} this method was
  first extended to embedded dynamics by Pikovsky
  ~\cite{pikovsky86-sov}, but more closely related to the prediction
  work of Sugihara \& May in \cite{sugihara90}}
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
prediction---provides accurate forecasts when the generating process
is a deterministic dynamical system.

Since LMA does not rest on an assumption of linearity (as ARIMA does),
it can handle both linear and nonlinear processes.  If the underlying
generating process is nondeterministic, however, it can perform
poorly.  In Figure~\ref{fig:embedding}b, for instance, little
structure is visible, so LMA may not work well.  More structure
appears to be present in~Figure~\ref{fig:embedding}c, but this
reconstruction also appears to contain some noise.  The question as to
how much structure is present in a reconstruction---and how much of
that structure can be captured and used by LMA---is apropos of the
central question treated in this paper.  It may be that \gcc has some
redundancy that LMA cannot exploit, or that the structure in \svdfive
is effectively obfuscated, from the standpoint of the LMA method, by
noise.  By quantifying the balance between redundancy (that is,
predictive structure) and entropy for these real-valued time series,
as shown in Section~\ref{sec:results}, we can begin to answer these
questions.

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 360
traces described in Section~\ref{sec:methods}, then calculate the
error between the predictions and the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, whose ability to generate accurate predictions is limited
only by the positive Lyapunov exponents of the system.  However, the
performance of the other three methods used here will degrade severely
if the associated models are not periodically rebuilt.  In order to
make the comparison fair, we used an iterative one-step prediction
schema \emph{for all four methods}.  This has the slightly confusing
effect of causing the ``test'' signal to be used both to assess the
accuracy of each model and for periodic refitting.

@@

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:svd5LMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  
% On this type of plot, a perfect prediction lies exactly on the
% diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error (MASE) between the true and predicted signals:
% 
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
% 
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  MASE is a normalized measure: the scaling
term in the denominator
% 
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
% 
is the average in-sample forecast error for a random-walk prediction
over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$ means that the
prediction error in question was, on the average, smaller than the
error of a random-walk forecast on the same data.  Analogously,
$MASE>1$ means that the corresponding prediction method did
\emph{worse}, on average, than the random-walk method.  We chose this
error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales.

### 2014-04-22 pecomplexity.tex
Liz is done rewriting

This paper provides insight into when, why, and how forecast
strategies fail when they are applied to complicated time series. We
conjecture that the inherent complexity of a time series —-- which
results from the inherent dimension, nonlinearity, and
non-stationarity of the generating process, as well as from
measurement issues like noise, aggregation, and finite data length ---
is both empirically quantifiable and directly correlated with
predictability. To validate this conjecture, we study 120 different
time-series data sets.  We argue that \emph{redundancy} is an
effective way to measure complexity and predictive structure in a time
series and that \emph{weighted permutation entropy} is an effective
way to estimate that complexity.  For each time series, we construct
predictions using a wide variety of forecast models, then compare the
accuracy of the predictions with the permutation entropy of that time
series.  We use the results to develop a heuristic that can help
practitioners recognize when a particular prediction method is not
well matched to the task at hand: that is, when the time series has
more predictive structure than that method can capture and exploit.

>>>>

This paper provides insight into when, why, and how forecast
strategies fail when they are applied to complicated time series. We
conjecture that the inherent complexity of a time series —-- which
results from the dimension, nonlinearity, and non-stationarity of the
generating process, as well as from measurement issues like noise,
aggregation, and finite data length --- is both empirically
quantifiable and directly correlated with predictability.  In
particular, we argue that \emph{redundancy} is an effective way to
measure complexity and predictive structure in a time series and that
\emph{weighted permutation entropy} is an effective way to estimate
that complexity.  To validate these conjectures, we study 120
different time-series data sets.  For each time series, we construct
predictions using a wide variety of forecast models, then compare the
accuracy of the predictions with the permutation entropy of that time
series.  We use the results to develop a heuristic that can help
practitioners recognize when a particular prediction method is not
well matched to the task at hand: that is, when the time series has
more predictive structure than that method can capture and exploit.

### 2014-04-22 related.tex
Liz is done rewriting

Modeling time-series data for the purposes of prediction dates back at
least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
then, hundreds, if not thousands, of strategies have been developed
for a wide variety of prediction tasks.  The purpose of this paper is
not to add a new weapon to this arsenal, nor to assess or compare the
effectiveness of existing methods.  Our goals are more general: {\sl
  (i)} to empirically quantify the predictive structure that is
present in a real-valued scalar time series and {\sl (ii)} to explore
how the performance of prediction methods is related to that inherent
complexity.  It would, of course, be neither practical nor interesting
to report results for every existing forecast method; instead, we
choose a representative set, as described in Section~\ref{sec:model}.

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most of the corresponding
solutions fall into two categories that we call model-based error
analysis and model-free information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
  used in different ways in the forecasting community.  Here, we
  distinguish those terms by the part of the time series that is the
  focus of the prediction: the observed data for the former and the
  unknown future for the latter.  In-sample forecasts---comparisons of
  predictions generated from \emph{part} of the observed time
  series---are useful for assessing model error and prediction
  horizons, among other things.}.
%
These distributions are then used as estimates of out-of-sample
forecast errors in those regions.  For example, Smith {\sl et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate forecast errors and verify that they are normally distributed.
If they are not, that indicates that there is structure in the time
series that the model-building process was unable to capture and use.
The problem with this approach is lack of generality.  
\label{page:normal-errors}
Normally distributed errors indicate that a model has captured the
structure in the data insofar as is possible, \emph{given the
  formulation of that particular model} (viz., the best possible
linear fit to a nonlinear dataset).  This gives no indication as to
whether another modeling strategy might do better.

@@

We sidestep these issues by using a variant of the \emph{permutation
  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate the value
of the Kolmogorov-Sinai entropy of a real-valued time series---and
thus the redundancy in that data, which is an effective proxy for
predictability.  This differs from existing approaches in a number of
ways.  It does not rely on generating partitions---and thus does not
introduce bias into the results if one does not know the dynamics or
cannot compute the partition.  It makes no assumptions about, and
requires no knowledge of, the underlying generating process: linear,
nonlinear, the Lyapunov spectrum, etc.  These features make our
approach applicable to noisy real-valued time series from all classes
of systems.

>>>>

Modeling time-series data for the purposes of prediction dates back at
least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
then, hundreds, if not thousands, of strategies have been developed
for a wide variety of prediction tasks.  The purpose of this paper is
not to add a new weapon to this arsenal, nor to assess or compare the
effectiveness of existing methods.  Our goals are more general:
\emph{(i)} to empirically quantify the predictive structure that is
present in a real-valued scalar time series and \emph{(ii)} to explore
how the performance of prediction methods is related to that inherent
complexity.  It would, of course, be neither practical nor interesting
to report results for every existing forecast method; instead, we
choose a representative set, as described in Section~\ref{sec:model}.

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most of the corresponding
solutions fall into two categories that we call model-based error
analysis and model-free information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
  used in different ways in the forecasting community.  Here, we
  distinguish those terms by the part of the time series that is the
  focus of the prediction: the observed data for the former and the
  unknown future for the latter.  In-sample forecasts---comparisons of
  predictions generated from \emph{part} of the observed time
  series---are useful for assessing model error and prediction
  horizons, among other things.}.
%
These distributions are then used as estimates of out-of-sample
forecast errors in those regions.  For example, Smith \emph{et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate forecast errors and verify that they are normally distributed.
If they are not, that suggests that there is structure in the time
series that the model-building process was unable to capture and use.
The problem with this approach is lack of generality.  
\label{page:normal-errors}
Normally distributed errors indicate that a model has captured the
structure in the data insofar as is possible, \emph{given the
  formulation of that particular model} (viz., the best possible
linear fit to a nonlinear dataset).  This gives no indication as to
whether another modeling strategy might do better.

@@

We sidestep these issues by using a variant of the \emph{permutation
  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate the value
of the Kolmogorov-Sinai entropy of a real-valued time series---and
thus the redundancy in that data, which we show to be an effective proxy for
predictability.  This differs from existing approaches in a number of
ways.  It does not rely on generating partitions---and thus does not
introduce bias into the results if one does not know the dynamics or
cannot compute the partition.  It makes no assumptions about, and
requires no knowledge of, the underlying generating process: linear,
nonlinear, the Lyapunov spectrum, etc.  These features make our
approach applicable to noisy real-valued time series from all classes
of systems.
